ExpressionView

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
EV: improved vignettes.

@@

\section{Problem}
The problem can be formulated very generally:
Given a matrix in which each element is part of at least one bicluster, arrange the rows and columns
in such a way that the biclusters are easily recognizable on a two-dimensional plot.
Since biclusters are rectangular, rows and columns can be ordered separately.
The problem thus reduces to finding the optimal arrangement of a set of $n$ elements 
that are part of at least one of $m$ clusters. In what follows, we refer to the $n$ elements as
slots.

@@

\section{Optimizing the order}
Optimizing the order thus consists in maximizing $Q$ by changing the order of the slots.
To achieve this goal we use a brute-force approach that 
\begin{itemize}
	\item{shifts parts of a given cluster to better positions and}
	\item{permutes elements contained in a given cluster.}
\end{itemize}

@@

The ordering is done in four steps:
\begin{enumerate}
	\item{\code{simplify}: Reduces the dimensions of the problem by excluding duplicate rows. The multiplicity of each slot is taken into account when calculating the score $Q$.}
	\item{\code{prearrange}: Tries to find a good starting point for the optimization by creating a new order, starting from the first row and adding the consecutive rows one after the other, placing them at the optimal position. For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
	\item{\code{arrange}: Uses the brute-force approach to further increase the alignment score.}
	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
\end{enumerate}

@@

The ordering iterates over the clusters and first calls the function \code{reposition}, taking \code{cluster} as argument. It goes through all the rows contained in a given cluster and tries to shift them to a better position. A contiguous group of rows can be repositioned at once or split up in subgroups. Empirically, it turned out that the alignment is fastest by starting the repositioning with a single row, increasing the length of the group only if no better position could be found. The opposite approach can be used by replacing the \emph{start forward} section on line 545 by the \emph{start backwards} code on line 556.

After repositioning, the main loop calls the exchange function, taking \code{cluster} and \code{what} as argument. It exchanges rows contained (\code{what = 1}) / not contained (\code{what = 0}) in a given cluster. It randomly picks \code{maxslots} number of rows (currently six), generates all possible permutations, calculates the corresponding alignment scores, and retains the best order. This realignment is repeated \code{maxiter} times, so that on average every slot is part of one optimization.

The loop over clusters calls the \code{reposition} function first for all the rows not contained in a given cluster and afterwards for the remaining ones. The main loop is repeated until the maximal computation time is reached or no improved alignment could be found anymore.

@@

\section{Possible improvements}
More than 90\% of the calculation time is spent to recalculate the length of the longest contiguous subclusters. Optimizing the \code{getfitness} routine
thus has a direct impact on the overall performance of the algorithm. Since the different clusters can be treated independently, it is
straightforward to parallelize the calculation, maybe even on the graphic chip.

>>>>

\setlength{\parindent}{2em}

@@

\section{Problem}
The problem can be formulated very generally:
given a matrix in which each element is part of at least one bicluster, arrange the rows and columns
in such a way that the biclusters are easily recognizable on a two-dimensional plot.
Since biclusters are rectangular, rows and columns can be ordered separately.
The problem thus reduces to finding the optimal arrangement of a set of $n$ elements 
that are part of at least one of $m$ clusters. In what follows, we refer to the $n$ elements as
slots.

@@

\section{Optimizing the order}
Optimizing the order thus consists in maximizing $Q$ by changing the order of the slots.
To achieve this goal we use a greedy approach that 
\begin{itemize}
	\item{shifts parts of a given cluster to better positions and}
	\item{permutes elements contained in a given cluster.}
\end{itemize}

@@

The ordering is done in four steps:
\begin{enumerate}
	\item{\code{simplify}: Reduces the dimensions of the problem by excluding duplicate rows. The multiplicity of each slot is taken into account when calculating the score $Q$.}
	\item{\code{prearrange}: Tries to find a good starting point for the optimization by creating a new order, starting from the first row and adding the consecutive rows one after the other, placing them at the optimal position. For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
	\item{\code{arrange}: Uses the greedy approach to further increase the alignment score.}
	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
\end{enumerate}

@@

The ordering iterates over the clusters and first tries to shift the
rows of them to a better position. A contiguous group of rows can be
repositioned at once or split up in subgroups. Empirically, it turned
out that the alignment is fastest by starting the repositioning with a
single row, increasing the length of the group only if no better
position could be found.

After repositioning, the algorithm tries to exchanges rows
contained or not contained in a given cluster. It randomly picks
a number of rows (up to six right now) and generates all their
possible permutations, calculates the corresponding alignment scores,
and retains the best order. This realignment is repeated
a number of times, so that on average every slot is part of one
optimization.

These two steps are repeated until the maximal computation time is
reached or no improved alignment could be found anymore.

@@

\section{Possible improvements}
More than 90\% of the calculation time is spent to recalculate the
length of the longest contiguous subclusters. Optimizing this
algorithm thus has a direct impact on the overall performance of the
algorithm. Since the different clusters can be treated independently,
it is straightforward to parallelize the calculation, maybe even on
the graphic chip.

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
ExpressionView paper and docs corrections.
EV: improved vignettes.

\section{Introduction}
Clustering genes according to their expression profiles is an important task 
in analyzing microarray data. In this tutorial, we explain how to use ExpressionView, 
an R package designed to interactively explore biclusters identified in gene 
expression data in conjunction with the Iterative Signature Algorithm (ISA)~\cite{bergmann03} 
and the biclustering methods available in the \Rpackage{Biclust} package~\cite{kaiser08}.

\section{Loading the gene expression data}
The \Rpackage{ExpressionView} package requires the gene expression data to be available in the
form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}. In this tutorial we will use the 
\Rpackage{Bioconductor} sample data from a 
clinical trial in acute lymphoblastic leukemia provided by the \Rpackage{ALL} package.
<<load the data>>=
library(ALL)
library(hgu95av2.db)
data(ALL)
@
The data set contains \Sexpr{ncol(ALL)} samples and \Sexpr{nrow(ALL)} features.

\section{Find biclusters}
There are many biclustering algorithms described in the literature~\cite{madeira04}.
All of them aim to reduce the complexity of the gene expression data by identifying 
suitable groups of genes and conditions that are co-regulated. In what follows we
show how to use two of the freely available biclustering methods.

@@

The ISA biclustering is not binary, every gene (and
similarly, every sample or condition) has a score between -1 and 1; the further the
score is from zero, the stronger the association between the gene
(or sample) and the module. If two features both have scores with the
same sign, they are correlated, otherwise, they are anti-correlated.

@@

The standard R function \Rfunction{as} 
can be used to convert \Rclass{ISAModules} to \Rclass{Biclust} objects. This
requires the binarization of the modules, i.e., the ISA scores are
lost since they are converted to zeros and ones:
<<biclust>>=
as(modules, "Biclust")
@

The conversion also works in the other direction. Starting from a \Rclass{Biclust} object, just use
<<isa>>=
as(biclusters, "ISAModules")
@
to get an \Rclass{ISAModules} object.

@@

\section{Order}
To present the tens of possibly overlapping biclusters in a visually appealing form, 
it is necessary to regroup the rows and columns of the gene expression matrix in 
such a way that biclusters form contiguous rectangles. Since for more than two 
mutually overlapping biclusters, it is in general impossible to find such an arrangement,
 one has to make concessions. In contrast to~\cite{grothaus06}, who proposed to repeat 
 rows and columns as necessary to achieve this goal, we prefer to optimize the 
 arrangement within the original data by maximizing the area of the largest contiguous biclusters.

The \Rfunction{OrderEV} function implemented in the \Rpackage{ExpressionView} package
determines the optimal order of the gene expression matrix for a given set of biclusters:
<<expressionview, results=HIDE>>=
library(ExpressionView)
optimalorder <- OrderEV(modules)
@

The result is a list containing various mappings between the original data and the optimal arrangement.
Note that the genes and the samples can be ordered separately. Apart form reordering the 
full gene expression matrix, the algorithm also determines the best arrangement of individual biclusters.
The mapping of the genes and the samples contained in bicluster \variable{i} can be accessed by 
<<gene maps, eval=FALSE>>=
optimalorder$genes[i+1]
optimalorder$samples[i+1]
@
Here, \variable{i=0}, returns the map of the complete data set. By default, the \Rfunction{OrderEV}
runs for roughly one minute, this might not be sufficient to find an appropriate order for data 
containing many overlapping biclusters. The status of the ordering is stored in 
<<status ordering>>=
optimalorder$status
@
If the status is set to \variable{1}, the algorithm has found the optimal solution. A \variable{0} indicates that the
the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} 
accepts two additional parameters to circumvent the problem of partial alignment:
One can start the ordering from a given initial configuration, i.e., the result of a previous arrangement
by defining the \Rargument{initialorder} argument
<<status ordering, eval=FALSE>>=
optimalorderp <- OrderEV(modules, initialorder=optimalorder, maxtime=120)
@
and one can increase the time limit by specifying \Rargument{maxtime}. Note that the time is indicated in seconds
and cannot be  smaller than 1.

\section{Export}
The \Rfunction{ExportEV} function allows the user to combine the available data and export it to an XML file that can be read by the Flash applet:
<<export, eval=FALSE>>=
ExportEV(modules, ALL, optimalorder, filename="file.evf")
@
The function gathers the data contained in the \Rclass{ExpressionSet} \variable{ALL}, orders it according to the optimal arrangement 
\variable{optimalorder} and adds the biclusters defined in \variable{modules}. The output is an uncompressed XML file that can be opened with any text viewer. We have chosen to use the extension \variable{.evf} (for ExpressionView file) for the data files. This extension is associated with the stand-alone version of the viewer, so that one can simply double-click on such a file to launch the program and load the data. The file association is the reason why we do not use the \variable{.xml} extension. A description of the XML layout can be found on the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. Before exporting the data, the 
\Rfunction{ExportEV} function automatically calculates GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the given biclusters.

@@

	\item[g, h] Modules are highlighted as the user moves the moves over the gene expression data. The two check boxes allow you to choose between highlighting all the parts of a module (Filling) or alternatively only the largest rectangle (Outline). You can also turn it off completely. For data sets with many modules, it can be helpful to restrict highlighting to Outline.

@@

The \Rfunction{ExportEV} uses the named list provided by the \variable{description} variable to label the data. First, let us annotate the rows and columns of the data set
<<data set annotation>>=
v <- vector("character"); for ( i in 1:dim(data)[1] ) { v[i] <- paste("row ", i, sep="") }; rownames(data) <- v
v <- vector("character"); for ( i in 1:dim(data)[2] ) { v[i] <- paste("column ", i, sep="") }; colnames(data) <- v
@

Next, we assign the meta data associated with the rows of the data matrix. In this example we use 5 tags labelled ''row tag``:
<<rowdata annotation>>=
rowdata <- matrix(nrow=dim(data)[1], ncol=5)
rownames(rowdata) <- rownames(data)
v <- vector("character"); for ( i in 1:dim(rowdata)[2] ) { v[i] <- paste("row tag ", i, sep="") }; colnames(rowdata) <- v
for ( i in 1:dim(rowdata)[1] ) { for ( j in 1:dim(rowdata)[2] ) { rowdata[i,j] <- paste("row description (", i, ", ", j, ")", sep="") }}
@
And similarly for the columns, using 10 "column tags``:
<<coldata annotation>>=
coldata <- matrix(nrow=dim(data)[2], ncol=10)
rownames(coldata) <- colnames(data)
v <- vector("character"); for ( i in 1:dim(coldata)[2] ) { v[i] <- paste("column tag ", i, sep="") }; colnames(coldata) <- v
for ( i in 1:dim(coldata)[1] ) { for ( j in 1:dim(coldata)[2] ) { coldata[i,j] <- paste("column description (", i, ", ", j, ")", sep="") }}
@

>>>>

\section{Introduction}
Clustering genes according to their expression profiles is an important task 
in analyzing microarray data. In this tutorial, we explain how to use ExpressionView, 
an R package designed to interactively explore biclusters identified in gene 
expression data, in conjunction with the Iterative Signature Algorithm (ISA)~\cite{bergmann03} 
and the biclustering methods available in the \Rpackage{Biclust} package~\cite{kaiser08}.

\section{Loading the gene expression data}
The \Rpackage{ExpressionView} package requires the gene expression data to be available in the
form of a BioConductor \Rclass{ExpressionSet}. In this tutorial we will use the 
BioConductor sample data from a 
clinical trial in acute lymphoblastic leukemia provided by the \Rpackage{ALL} package.
<<load the data>>=
library(ALL)
library(hgu95av2.db)
data(ALL)
@
The data set contains \Sexpr{ncol(ALL)} samples and \Sexpr{nrow(ALL)} features.

\section{Find biclusters}
There are many biclustering algorithms described in the literature~\cite{madeira04}.
All of them aim to reduce the complexity of the gene expression data by identifying 
suitable groups of genes and conditions that are co-expressed. In this tutorial
show how to use \Rpackage{ExpressionView} with some of the available biclustering algorithms.

@@

This object can be directly used with the functions of the
\Rpackage{ExpressionView} package. See Section~\ref{sec:ordering} for
details.

@@

\Rclass{Biclust} objects can be directly used with the
\Rpackage{ExpressionView} functions. Alternatively, they can be
converted to \Rclass{ISAModules} objects, using the standard
\Rfunction{as} R function:
<<isa>>=
as(biclusters, "ISAModules")
@
results an \Rclass{ISAModules} object.

@@

\section{Order}%
\label{sec:ordering}
To present the tens of possibly overlapping biclusters in a visually appealing form, 
it is necessary to reorder the rows and columns of the gene expression matrix in 
such a way that biclusters form contiguous rectangles. Since for more than two 
mutually overlapping biclusters, it is in general impossible to find such an arrangement,
 one has to make concessions. In contrast methods that propose to repeat 
 rows and columns as necessary to achieve this goal~\cite{grothaus06}, we prefer to optimize the 
 arrangement within the original data by maximizing the area of the largest contiguous biclusters.

The \Rfunction{OrderEV} function implemented in the \Rpackage{ExpressionView} package
determines the optimal order of the gene expression matrix for a given
set of biclusters. It can be called with \Rclass{ISAModules} or
\Rclass{Biclust} objects as the first argument:
<<expressionview, results=HIDE>>=
library(ExpressionView)
optimalorder <- OrderEV(modules)
@

The result is a list containing various mappings between the original data and the optimal arrangement.
Note that the genes and the samples can be ordered separately. Apart form reordering the 
full gene expression matrix, the algorithm also determines the best arrangement of individual biclusters.
The mapping of the genes and the samples contained in bicluster \variable{i} can be accessed by 
<<gene maps, eval=FALSE>>=
optimalorder$genes[i+1]
optimalorder$samples[i+1]
@
The first elements of the lists contain the optimal ordering of the
complete matrix. By default, the \Rfunction{OrderEV} function
runs for roughly one minute, this might not be sufficient to find an appropriate order for data 
containing many overlapping biclusters. The status of the ordering is stored in 
<<status ordering>>=
optimalorder$status
@
If the status is set to \variable{1}, the algorithm has found the optimal solution. A \variable{0} indicates that the
the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} function
accepts two additional parameters to circumvent the problem of partial alignment:
One can start the ordering from a given initial configuration, i.e., the result of a previous arrangement
by defining the \Rargument{initialorder} argument
<<status ordering, eval=FALSE>>=
optimalorderp <- OrderEV(modules, initialorder=optimalorder, maxtime=120)
@
and one can increase the time limit by specifying \Rargument{maxtime}. Note that the time is indicated in seconds
and cannot be  smaller than 1.

\section{Export}
The \Rfunction{ExportEV} function allows the user to combine the available data and export it to an XML file that can be read by the Flash applet:
<<export, eval=FALSE>>=
ExportEV(modules, ALL, optimalorder, filename="file.evf")
@
The function gathers the data contained in the \Rclass{ExpressionSet} \variable{ALL}, orders it according to the optimal arrangement 
\variable{optimalorder} and adds the biclusters defined in \variable{modules}. The output is an uncompressed XML file that can be opened with any text viewer. We have chosen to use the extension \variable{.evf} (for ExpressionView file) for the data files. This extension is associated with the stand-alone version of the viewer, so that one can simply double-click on such a file to launch the program and load the data. The file association is the reason why we do not use the \variable{.xml} extension. A description of the XML layout can be found on the \Rpackage{ExpressionView} website at \url{http://www.unil.ch/cbg/ExpressionView}. Before exporting the data, the
\Rfunction{ExportEV} function automatically calculates GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the given biclusters.

@@

	\item[g, h] Modules are highlighted as the user moves the mouse over the gene expression data. The two check boxes allow you to choose between highlighting all the parts of a module (Filling) or alternatively only the largest rectangle (Outline). You can also turn it off completely. For data sets with many modules, it can be helpful to restrict highlighting to Outline.

@@

The \Rfunction{ExportEV} uses the named list provided by the \variable{description} variable to label the data. First, let us annotate the rows and columns of the data set
<<data set annotation>>=
rownames(data) <- paste("row", seq_len(nrow(data)))
colnames(data) <- paste("column", seq_len(ncol(data)))
@

Next, we assign the meta data associated with the rows of the data matrix. In this example we use 5 tags labelled ''row tag``:
<<rowdata annotation>>=
rowdata <- outer(1:nrow(data), 1:5, function(x, y) {
  paste("row description (", x, ", ", y, ")", sep="")
})             
rownames(rowdata) <- rownames(data)
colnames(rowdata) <- paste("row tag", seq_len(ncol(rowdata)))
@
And similarly for the columns, using 10 ``column tags'':
<<coldata annotation>>=
coldata <- outer(1:ncol(data), 1:10, function(x, y) {
  paste("column description (", x, ", ", y, ")", sep="")
})
rownames(coldata) <- colnames(data)
colnames(coldata) <- paste("column tag", seq_len(ncol(coldata)))
@
